---
title: "Getting and cleaning data"
author: "Sanidhya"
date: "2026-03-23"
output: html_document
---

````markdown
---
title: "Reproducible Research Assignment 1"
author: "Your Name"
output: html_document
---

## Loading and preprocessing the data

```{r, echo=TRUE}
library(dplyr)
library(lattice)

data <- read.csv("activity.csv")
data$date <- as.Date(data$date)
```

---

## What is mean total number of steps taken per day?

```{r, echo=TRUE}
steps_per_day <- data %>%
  group_by(date) %>%
  summarise(total_steps = sum(steps, na.rm = TRUE))
```

### Histogram

```{r, echo=TRUE}
hist(steps_per_day$total_steps,
     main = "Total Steps per Day",
     xlab = "Steps",
     col = "blue")
```

### Mean and Median

```{r, echo=TRUE}
mean_steps <- mean(steps_per_day$total_steps)
median_steps <- median(steps_per_day$total_steps)

mean_steps
median_steps
```

**Observation:**
The mean and median represent the average daily steps ignoring missing values.

---

## What is the average daily activity pattern?

```{r, echo=TRUE}
interval_avg <- data %>%
  group_by(interval) %>%
  summarise(avg_steps = mean(steps, na.rm = TRUE))
```

### Time Series Plot

```{r, echo=TRUE}
plot(interval_avg$interval, interval_avg$avg_steps,
     type = "l",
     xlab = "Interval",
     ylab = "Average Steps")
```

### Maximum Interval

```{r, echo=TRUE}
interval_avg[which.max(interval_avg$avg_steps), ]
```

**Observation:**
This interval has the highest average activity.

---

## Imputing missing values

### Total missing values

```{r, echo=TRUE}
sum(is.na(data$steps))
```

### Strategy

Missing values are replaced using the **mean of the respective 5-minute interval**.

```{r, echo=TRUE}
interval_means <- data %>%
  group_by(interval) %>%
  summarise(mean_steps = mean(steps, na.rm = TRUE))

data_filled <- data

data_filled$steps[is.na(data_filled$steps)] <-
  interval_means$mean_steps[
    match(data_filled$interval[is.na(data_filled$steps)],
          interval_means$interval)
  ]
```

---

## Histogram after imputing missing values

```{r, echo=TRUE}
steps_per_day_filled <- data_filled %>%
  group_by(date) %>%
  summarise(total_steps = sum(steps))

hist(steps_per_day_filled$total_steps,
     main = "Total Steps per Day (After Imputation)",
     xlab = "Steps",
     col = "green")
```

### Mean and Median after filling

```{r, echo=TRUE}
mean(steps_per_day_filled$total_steps)
median(steps_per_day_filled$total_steps)
```

**Observation:**
After imputing missing values, the mean increases slightly. Missing data earlier caused underestimation.

---

## Are there differences in activity patterns between weekdays and weekends?

```{r, echo=TRUE}
data_filled$day_type <- ifelse(
  weekdays(data_filled$date) %in% c("Saturday", "Sunday"),
  "weekend",
  "weekday"
)
```

### Average steps by interval and type

```{r, echo=TRUE}
avg_by_daytype <- data_filled %>%
  group_by(interval, day_type) %>%
  summarise(avg_steps = mean(steps))
```

### Panel Plot

```{r, echo=TRUE}
xyplot(avg_steps ~ interval | day_type,
       data = avg_by_daytype,
       type = "l",
       layout = c(1,2),
       xlab = "Interval",
       ylab = "Average Steps")
```

**Observation:**
Weekday activity is more structured, while weekend activity is more spread out.

```
```

